How topos theory can help algebra and geometry

### 2018-07-09 slides-leipzig2018.tex
Slides for the Leipzig talk finished

      \includegraphics[height=0.35\textwidth]{grothendieck}
      \quad
      \includegraphics[height=0.35\textwidth]{elliptic-curve}

      étale topos of a scheme
    \end{column}

@@

      \begin{tikzpicture}
        \node (overview) {
          \scalebox{0.32}{\sieve{6}{4}}
        };
        \def\R{8pt}
        \begin{pgfonlayer}{background}
        \draw[decoration={bumps,segment length=8pt}, decorate, very thick, draw=mypurple]
          ($(overview.south west) + (\R, 0)$) arc(270:180:\R) --
          ($(overview.north west) + (0, -\R)$) arc(180:90:\R) --
          ($(overview.north east) + (-\R, 0)$) arc(90:0:\R) --
          ($(overview.south east) + (0, \R)$) arc(0:-90:\R) --
          cycle;
        \end{pgfonlayer}
      \end{tikzpicture}
    \end{column}
  \end{columns}

@@

      \includegraphics[height=0.35\textwidth]{sheaf}

@@

      \includegraphics[height=0.35\textwidth]{olivia-lattices}

@@

>>>>

      \includegraphics[height=0.32\textwidth]{grothendieck}
      \quad
      \includegraphics[height=0.32\textwidth]{elliptic-curve}

      \footnotesize
      étale topos of a scheme \\
      field with one element
    \end{column}

@@

      \begin{tikzpicture}
        \node (overview) {
          \scalebox{0.44}{\sieve{6}{4}}
        };
        \def\R{8pt}
        \begin{pgfonlayer}{background}
        \draw[decoration={bumps,segment length=8pt}, decorate, very thick, draw=mypurple]
          ($(overview.south west) + (\R, 0)$) arc(270:180:\R) --
          ($(overview.north west) + (0, -\R)$) arc(180:90:\R) --
          ($(overview.north east) + (-\R, 0)$) arc(90:0:\R) --
          ($(overview.south east) + (0, \R)$) arc(0:-90:\R) --
          cycle;
        \end{pgfonlayer}
      \end{tikzpicture}
    \end{column}
  \end{columns}

@@

      \includegraphics[height=0.32\textwidth]{sheaf}

@@

      \includegraphics[height=0.32\textwidth]{olivia-lattices}

@@

\newcommand{\appl}[3]{\centering\includegraphics[height=3.5em]{#1}\par\hil{#2}\\\footnotesize#3}

\begin{frame}{Applications}
  \bigskip

  \small
  \begin{columns}
    \begin{column}{0.35\textwidth}
      \appl{analysis}{analysis}{
        internalizing parameter-dependence
      }
    \end{column}

    \begin{column}{0.39\textwidth}
      \appl{calabi-yau}{algebraic geometry}{
        \mbox{reducing geometry to algebra} \\
        \mbox{reducing relative to absolute} \\
        synthetic account
      }
    \end{column}

    \begin{column}{0.35\textwidth}
      \appl{klein-bottle}{differential geometry}{
        reflection principles \\
        synthetic account
      }
    \end{column}
  \end{columns}

  \begin{columns}
    \begin{column}{0.35\textwidth}
      \appl{torus-rainbow}{homotopy theory}{
        synthetic account
        computer-assisted proofs
        generalizations
      }
    \end{column}

    \begin{column}{0.39\textwidth}
      \appl{generic-freeness}{commutative algebra}{
        local-to-global principles \\
        reduction techniques \\
        constructive proofs
      }
    \end{column}

    \begin{column}{0.35\textwidth}
      \appl{bohr-topos}{further subjects}{
        synth.~computability th. \\
        synth.~measure theory \\
        Bohr topos for QM
      }
    \end{column}
  \end{columns}
\end{frame}

\section[Analysis]{Applications in analysis}

\begin{frame}{The topos of sheaves over a space}
  \small\vspace*{-0.8em}
  Let~$X$ be a topological space. We recursively define
  \[ U \models \varphi \quad \text{(``$\varphi$ holds on~$U$'')} \]
  for open subsets~$U \subseteq X$ and formulas~$\varphi$.
  Write~``$\Sh(X) \models \varphi$'' to mean~$X \models \varphi$.
  Let~$\C(U)$ be the set of continuous functions~$U \to \RR$.

  \vspace*{-1.3em}
  \footnotesize
  \[ \renewcommand{\arraystretch}{1.15}\begin{array}{@{}l@{\ \ }c@{\ \ }l@{}}
  U \models \top &\Ll& \text{true} \\
  U \models \bot &\Ll& \hcancel{\text{false}}{0pt}{3pt}{0pt}{-2pt}\ U = \emptyset \\
  U \models s = t &\Ll& \text{$s(x) = t(x)$ for all $x \in U$} \\
  U \models \varphi \wedge \psi &\Ll&
  \text{$U \models \varphi$ and $U \models \psi$} \\
  U \models \varphi \vee \psi &\Ll&
  \hcancel{\text{$U \models \varphi$ or $U \models \psi$}}{0pt}{3pt}{0pt}{-2pt}\ \text{there exists a covering $U = \bigcup_i U_i$} \\
  && \quad\quad\text{such that for all~$i$: $U_i \models \varphi$ or $U_i \models \psi$} \\
  U \models \varphi \Rightarrow \psi &\Ll&
  \text{for all open~$V \subseteq U$: }
  \text{$V \models \varphi$ implies $V \models \psi$} \\
  U \models \forall s \? \RR\_ \varphi(s) &\Ll&
  \text{for all open $V \subseteq U$ and functions~$s_0 \in \C(V)$: $V \models \varphi(s_0)$} \\
  U \models \exists s \? \RR\_ \varphi(s) &\Ll&
  \hcancel{\text{there exists $s_0 \in \C(U)$ such that $U \models \varphi(s_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
  &&
  \text{there exists a covering $U = \bigcup_i U_i$ such that for all~$i$:} \\
  && \quad\quad \text{there exists~$s_0 \in \C(U_i)$ such that $U_i \models \varphi(s_0)$}
  \end{array} \]
\end{frame}

\section[Differential geometry]{Applications in differential geometry}

\begin{frame}{Synthetic differential geometry}
  \vspace*{-1em}
  \begin{center}
    \includegraphics[width=0.5\textwidth]{microaffinity}
  \end{center}

  \vspace*{-2em}
  \begin{varblock}{10.8cm}{The axiom of microaffinity}
    \justifying
    Let~$\Delta = \{ \varepsilon \in \RR \,|\, \varepsilon^2 = 0 \}$.
    For any function $f : \Delta \to \RR$,
    there is a unique number~$a \in \RR$ such that
    $f(\varepsilon) = f(0) + a \varepsilon$
    for all~$\varepsilon \in \Delta$.
  \end{varblock}

  \vspace*{-1em}
  \small
  \begin{itemize}
    \item The \hil{derivative} of a function~$f : \RR \to \RR$ at~$x_0 \in \RR$
    is the unique number~$a \in \RR$ such that~$f(x_0 + \varepsilon) = f(x_0) +
    a \varepsilon$ for all~$\varepsilon \in \Delta$. \\[-1.2em]
    \item Manifolds are \hil{just sets}. \\[-1.2em]
    \item A \hil{tangent vector} to~$M$ is a map~$\Delta \to M$.
  \end{itemize}
\end{frame}

\section[Commutative algebra]{Applications to commutative algebra}

\begin{frame}{The little Zariski topos of a ring}
  \small\vspace*{-0.8em}
  Let~$A$ be a ring (commutative, with unit). We recursively define
  \[ D(f) \models \varphi \quad \text{(``$\varphi$ holds away from the zeros of~$f$'')} \]
  for~$f \in A$ and formulas~$\varphi$. Write
  ``$\Spec(A) \models \varphi$'' to mean~$D(1) \models \varphi$.

  \vspace*{-1em}
  \[\footnotesize\renewcommand{\arraystretch}{1.25}\begin{array}{@{}l@{\ }c@{\ }l@{}}
    D(f) \models \top &\text{iff}& \text{true} \\
    D(f) \models \bot &\text{iff}& \text{$f$ is nilpotent} \\
    D(f) \models x = y &\text{iff}& x = y \in A[f^{-1}] \\
    D(f) \models \varphi \wedge \psi &\text{iff}&
      \text{$D(f) \models \varphi$ and $D(f) \models \psi$} \\
    D(f) \models \varphi \vee \psi &\text{iff}&
      \text{there exists a partition~$f^n = fg_1 + \cdots + fg_m$ with,} \\
    &&\quad\text{for each~$i$, $D(fg_i) \models \varphi$ or $D(fg_i) \models \psi$} \\
    D(f) \models \varphi \Rightarrow \psi &\text{iff}&
      \text{for all~$g \in A$, $D(fg) \models \varphi$ implies $D(fg) \models \psi$} \\
    D(f) \models \forall x\?M^\sim\_ \varphi(x) &\text{iff}&
      \text{for all~$g \in A$ and $x_0 \in M[(fg)^{-1}]$, $D(fg) \models \varphi[x_0/x]$} \\
    D(f) \models \exists x\?M^\sim\_ \varphi(x) &\text{iff}&
      \text{there exists a partition~$f^n = fg_1 + \cdots + fg_m$ with,} \\
    &&\quad\text{for each~$i$, $D(fg_i) \models \varphi[x_0/x]$ for some~$x_0 \in M[(fg_i)^{-1}]$}
  \end{array} \]
\end{frame}

\section[Algebraic geometry]{Applications in algebraic geometry}

\begin{frame}{Understanding algebraic geometry}
  \vspace*{-1.5em}\small
  \begin{varblock}{0.9\textwidth}{}
    \justifying
    Understand \hil{notions of algebraic geometry} over a scheme~$X$ as
    \hil{notions of algebra} internal to~$\Sh(X)$.
  \end{varblock}

  \only<1>{
    \centering
    \rotatebox{90}{\tiny\scalebox{0.8}{Illustration: Carina Willbold}}\hspace{-0.05cm}%
    \includegraphics[width=0.7\textwidth]{external-internal-small}
    \par
  }

  \pause

  \small\centering
  \scalebox{0.83}{\begin{tabular}{ll}
    \toprule
    externally & internally to $\Sh(X)$ \\
    \midrule
    sheaf of sets & set \\
    % sheaf of rings & ring \\
    sheaf of modules & module \\
    % sheaf of finite type & finitely generated module \\
    finite locally free sheaf & finite free module \\
    % coherent sheaf & coherent module \\
    tensor product of sheaves & tensor product of modules \\
    % sheaf of Kähler differentials & module of Kähler differentials \\
    sheaf of rational functions & total quotient ring of~$\O_X$ \\
    dimension of $X$ & Krull dimension of~$\O_X$ \\
    spectrum of a sheaf of~$\O_X$-algebras & ordinary spectrum [with a twist] \\
    % big Zariski topos of $X$ & big Zariski topos of the ring $\O_X$ [with a twist] \\
    higher direct image & sheaf cohomology \\
    \bottomrule
  \end{tabular}}

  \footnotesize

  \begin{columns}[c]
    \begin{column}{0.47\textwidth}
      \begin{exampleblock}{}
        \justifying
        Let $0 \to \F' \to \F \to \F'' \to 0$ be a short exact sequence
        of sheaves of~$\O_X$-modules. If~$\F'$ and~$\F''$ are of finite type,
        so is~$\F$.
      \end{exampleblock}
    \end{column}

    \begin{column}{0.1\textwidth}
      \vspace*{0.7em}
      \scalebox{3}{$\Leftarrow$}
    \end{column}

    \begin{column}{0.44\textwidth}
      \begin{exampleblock}{}
        \justifying
        Let~$0 \to M' \to M \to M'' \to 0$ be a short exact sequence of
        modules. If~$M'$ and~$M''$ are finitely generated, so is~$M$.
      \end{exampleblock}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{Synthetic algebraic geometry}
  Usual approach to algebraic geometry: \hil{layer schemes above ordinary set theory}
  using either
  \begin{itemize}
    \item locally ringed spaces
    \small
    \begin{multline*}
      \text{set of prime ideals of~$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$} + {} \\
      \text{Zariski topology} + \text{structure sheaf}
    \end{multline*}
    \normalsize
    \item or Grothendieck's functor-of-points account, where a scheme is a
    functor~$\Ring \to \Set$.
    \small\[ A \longmapsto \{ (x,y,z) \in A^3 \,|\, x^n+y^n-z^n=0 \} \]
  \end{itemize}
  \bigskip

  \hil{Synthetic approach:} model schemes \hil{directly as sets} in
  the internal universe of the \mbox{\hil{big Zariski topos}} of a base scheme.
  \small
  \[ \{ (x,y,z) \? (\affl)^3 \,|\, x^n+y^n-z^n=0 \} \]
\end{frame}
